---
title: "Slides!"
author: "Dr. Shoemaker"
output:
  slidy_presentation: 
    highlight: kate
  powerpoint_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is this? 

in the rmd file, look at the YAML, you'll see that the output is listed as `slidy_presentation`. This is the output option that makes the html slide deck that you've seen in class. 

The drop down arrow next to knit shows you the different options you have. 

# R Markdown {.smaller}

```{r}
library(palmerpenguins)
x <- "x"
```

There are `r nrow(penguins)` in the dataset


This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

# Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

# Slide with R Output

Each slides starts when you add a new heading or subheading. 
```{r cars, echo = TRUE}
summary(cars)
```

# See? New Slide.

You can also manually break slides with three dashes

---

this makes a new slide too, one that doesn't have a heading. 

# Slide with Plot


```{r pressure}
plot(pressure)
```

